---
title: "Getting Started"
vignette: >
  %\VignetteIndexEntry{Getting Started}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
---

```{r}
#| include: false
#| eval: false
library(freestiler)
```

## Overview

freestiler is a Rust-powered vector tile engine for R and Python. It takes sf
data frames or GeoDataFrames and produces
[PMTiles](https://github.com/protomaps/PMTiles) archives -- self-contained,
single-file tilesets that can be served statically or viewed locally. The entire
pipeline runs in-memory with zero external dependencies: no tippecanoe, no Java,
no Go.

freestiler supports two tile formats:

- **MLT (MapLibre Tiles)** -- the default. A next-generation columnar binary
  format for efficient vector tile encoding.
- **MVT (Mapbox Vector Tiles)** -- the widely supported protobuf-based format.

## Installation

### R

Install from [r-universe](https://walkerke.r-universe.dev):

```r
install.packages("freestiler", repos = "https://walkerke.r-universe.dev")
```

The `r-universe` build is the recommended install for serious DuckDB-backed
workloads. Native builds include the Rust DuckDB backend by default, including
the streaming point pipeline used by `freestile_query()`.

Or install the development version from GitHub:

```r
# install.packages("devtools")
devtools::install_github("walkerke/freestiler")
```

### Python

Install from source (requires Rust toolchain):

```bash
cd python
pip install -e .
```

See the [Python Setup](python.html) article for detailed installation
instructions, virtual environment setup, and building with optional features.

## Creating your first tileset

The main function is `freestile()`. Pass it spatial data and an output path.

**R**

```{r}
#| eval: false
library(sf)
library(freestiler)

nc <- st_read(system.file("shape/nc.shp", package = "sf"))

freestile(nc, "nc_counties.pmtiles", layer_name = "counties")
```

```
Creating MLT tiles (zoom 0-14) for 100 features across 1 layer...
  Tiling layer 'counties' (zoom 0-14)...
Created nc_counties.pmtiles (65.2 KB)
```

**Python**

```python
import geopandas as gpd
from freestiler import freestile

gdf = gpd.read_file("nc.shp")

freestile(gdf, "nc_counties.pmtiles", layer_name="counties")
```

```
Creating MLT tiles (zoom 0-14) for 100 features across 1 layer...
  Tiling layer 'counties' (zoom 0-14)...
Created nc_counties.pmtiles (65.2 KB)
```

## Viewing tiles

### R

Use the [mapgl](https://walker-data.com/mapgl/) package to view your tileset
in R. mapgl supports both MLT and MVT tiles. PMTiles need HTTP range requests,
so start a local file server first:

```bash
# Serve from /tmp on port 8082 (pick whichever you prefer)
npx http-server /tmp -p 8082 --cors -c-1
```

Then point mapgl at the URL:

```{r}
#| eval: false
library(mapgl)

maplibre() |>
  add_pmtiles_source(
    id = "counties-src",
    url = "http://localhost:8082/nc_counties.pmtiles"
  ) |>
  add_fill_layer(
    id = "county-fill",
    source = "counties-src",
    source_layer = "counties",
    fill_color = "#00897b",
    fill_opacity = 0.5
  ) |>
  add_line_layer(
    id = "county-borders",
    source = "counties-src",
    source_layer = "counties",
    line_color = "#004d40",
    line_width = 1
  )
```

### Python

Python viewers like py-maplibregl and leafmap currently bundle MapLibre GL JS
versions earlier than 5.17, which means they support MVT but not yet MLT. For
Python viewing, create tiles with `tile_format="mvt"`:

```python
freestile(gdf, "nc_mvt.pmtiles", layer_name="counties", tile_format="mvt")
```

Then view with py-maplibregl:

```python
from maplibre import Map, MapOptions
from maplibre.sources import VectorTileSource
from maplibre.layers import Layer, LayerType

source = VectorTileSource(url="pmtiles://nc_mvt.pmtiles")

layer = Layer(
    type=LayerType.FILL,
    id="counties",
    source="counties-source",
    source_layer="counties",
    paint={"fill-color": "#00897b", "fill-opacity": 0.5}
)

m = Map(MapOptions(center=[-79.5, 35.5], zoom=6))
m.add_source("counties-source", source)
m.add_layer(layer)
m
```

MLT viewing in Python will work once these viewers upgrade to MapLibre GL JS
5.17+. In the meantime, you can view MLT tiles in the browser with
[MapLibre GL JS](https://maplibre.org/) directly.

## MLT vs MVT format

The default format is MLT (MapLibre Tiles). To use MVT instead:

**R**

```{r}
#| eval: false
freestile(nc, "nc_mvt.pmtiles", layer_name = "counties", tile_format = "mvt")
```

**Python**

```python
freestile(gdf, "nc_mvt.pmtiles", layer_name="counties", tile_format="mvt")
```

MLT is a columnar encoding that can be more compact than MVT, especially for
polygon-heavy datasets. MVT is the universal fallback supported by all mapping
libraries. Use MVT when maximum viewer compatibility is needed -- particularly
for Python viewers today.

## Controlling zoom levels

The `min_zoom` and `max_zoom` parameters control the zoom range.

**R**

```{r}
#| eval: false
freestile(nc, "nc_z4_10.pmtiles",
  layer_name = "counties",
  min_zoom = 4,
  max_zoom = 10
)
```

**Python**

```python
freestile(gdf, "nc_z4_10.pmtiles",
    layer_name="counties",
    min_zoom=4,
    max_zoom=10
)
```

## Feature dropping with drop_rate

For large datasets, `drop_rate` provides exponential feature thinning at lower
zoom levels. Points are thinned using spatial ordering (Morton curve) to
maintain even coverage; polygons and lines are thinned by area.

The `base_zoom` parameter controls the zoom level above which all features are
present. Below `base_zoom`, features are progressively thinned. By default,
`base_zoom` equals `max_zoom`.

**R**

```{r}
#| eval: false
freestile(nc, "nc_dropping.pmtiles",
  layer_name = "counties",
  drop_rate = 2.5,
  base_zoom = 10
)
```

**Python**

```python
freestile(gdf, "nc_dropping.pmtiles",
    layer_name="counties",
    drop_rate=2.5,
    base_zoom=10
)
```

## Direct file input

Tile spatial files directly without loading them into R or Python first.

**R** -- supports GeoParquet, GeoPackage, Shapefile, and other formats via DuckDB:

```{r}
#| eval: false
# GeoParquet (uses the geoparquet engine)
freestile_file("census_blocks.parquet", "blocks.pmtiles")

# GeoPackage, Shapefile, or other formats (uses the DuckDB engine)
freestile_file("counties.gpkg", "counties.pmtiles", engine = "duckdb")
```

**Python** -- GeoParquet works in the published wheels. GeoPackage,
Shapefile, and other DuckDB-backed formats require a source build with
DuckDB enabled:

```python
from freestiler import freestile_file

# GeoParquet (uses the geoparquet engine)
freestile_file("census_blocks.parquet", "blocks.pmtiles")

# GeoPackage, Shapefile, or other formats (uses the DuckDB engine)
freestile_file("counties.gpkg", "counties.pmtiles", engine="duckdb")
```

## SQL queries with DuckDB

Run a SQL query through DuckDB's spatial extension and pipe the results directly
into the tiling engine. Filter, join, and transform your data with SQL before
tiling.

**R**

```{r}
#| eval: false
freestile_query(
  "SELECT * FROM ST_Read('counties.shp') WHERE pop > 50000",
  "large_counties.pmtiles"
)

freestile_query(
  "SELECT * FROM read_parquet('blocks.parquet') WHERE state = 'NC'",
  "nc_blocks.pmtiles"
)
```

**Python** -- requires a source build with DuckDB enabled:

```python
from freestiler import freestile_query

freestile_query(
    "SELECT * FROM ST_Read('counties.shp') WHERE pop > 50000",
    "large_counties.pmtiles"
)

freestile_query(
    "SELECT * FROM read_parquet('blocks.parquet') WHERE state = 'NC'",
    "nc_blocks.pmtiles"
)
```

## Multi-layer tiles

Pass a named list (R) or dictionary (Python) to create multi-layer tilesets.

**R**

```{r}
#| eval: false
pts <- st_centroid(nc)

freestile(
  list(counties = nc, centroids = pts),
  "nc_layers.pmtiles"
)
```

Use `freestile_layer()` for per-layer zoom control:

```{r}
#| eval: false
freestile(
  list(
    counties = freestile_layer(nc, min_zoom = 0, max_zoom = 10),
    centroids = freestile_layer(pts, min_zoom = 6, max_zoom = 14)
  ),
  "nc_layers.pmtiles"
)
```

**Python** -- use `freestile_layer()` for per-layer zoom control:

```python
from freestiler import freestile, freestile_layer

centroids = gdf.copy()
centroids.geometry = gdf.geometry.centroid

freestile(
    {
        "counties": freestile_layer(gdf, min_zoom=0, max_zoom=10),
        "centroids": freestile_layer(centroids, min_zoom=6, max_zoom=14),
    },
    "nc_layers.pmtiles"
)
```

## Point clustering

For point layers, `cluster_distance` merges nearby points into clusters with
a `point_count` attribute.

**R**

```{r}
#| eval: false
freestile(pts, "nc_clustered.pmtiles",
  layer_name = "centroids",
  cluster_distance = 50,
  cluster_maxzoom = 8
)
```

**Python**

```python
freestile(centroids, "nc_clustered.pmtiles",
    layer_name="centroids",
    cluster_distance=50,
    cluster_maxzoom=8
)
```

## Feature coalescing

The `coalesce` parameter merges features with identical attributes within each
tile. Lines sharing endpoints are joined; polygons are grouped into
MultiPolygons.

**R**

```{r}
#| eval: false
freestile(nc, "nc_coalesced.pmtiles",
  layer_name = "counties",
  coalesce = TRUE
)
```

**Python**

```python
freestile(gdf, "nc_coalesced.pmtiles",
    layer_name="counties",
    coalesce=True
)
```
