---
title: "MapLibre Tiles (MLT)"
vignette: >
  %\VignetteIndexEntry{MapLibre Tiles (MLT)}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
---

```{r}
#| include: false
#| eval: false
library(freestiler)
```

freestiler supports the [MapLibre Tiles (MLT)](https://github.com/maplibre/maplibre-tile-spec) format, a columnar binary tile encoding announced in January 2026 by the MapLibre organization. While the default format is MVT for broad compatibility, MLT can produce smaller files for polygon and line data. This article explains what MLT is, how it compares to MVT, and when you might want to use one vs the other.

### Vector tiles in brief

Vector tiles divide geographic data into a grid of square tiles at multiple zoom levels. Unlike raster tiles (pre-rendered images), vector tiles store the actual geometry and attribute data, so the client can style, interact with, and dynamically render the data. The standard container for vector tilesets is [PMTiles](https://github.com/protomaps/PMTiles) - a single-file archive you can serve from any static file host.

### MVT: the standard

[Mapbox Vector Tiles (MVT)](https://github.com/mapbox/vector-tile-spec) has been the dominant vector tile encoding since Mapbox introduced it in 2014. It uses Protocol Buffers to serialize geometry and attributes, and it's supported by every major mapping library: MapLibre GL, Mapbox GL, Leaflet, deck.gl, and more. When you need maximum compatibility, MVT is the safe choice.

### MLT: what's different

Where MVT stores data row-by-row (one feature at a time), MLT organizes data column-by-column - all x-coordinates together, all y-coordinates together, all values of a given attribute together. This columnar layout opens up compression techniques that aren't available in MVT's row-oriented format: delta encoding for coordinates, run-length encoding for repeated values, and dictionary encoding for string columns with low cardinality.

The practical result is smaller tiles, especially for polygon-heavy datasets:

| Dataset | MVT | MLT | Savings |
|---------|-----|-----|---------|
| NC counties (z0-14) | 78 KB | 65 KB | 17% |
| US block groups (z4-12) | 310 MB | 259 MB | 16% |
| Education dots (z0-12) | 238 MB | 237 MB | ~0% |

Point-only datasets see minimal savings because coordinate data dominates and compresses similarly in both formats.

### Which format should I use?

Use **MLT** when you're working in R with [mapgl](https://walker-data.com/mapgl/) or viewing in a browser with MapLibre GL JS 5.21+, and you want smaller file sizes for polygon and line data.

Use **MVT** (the default) for the widest viewer compatibility - it works with MapLibre GL JS, Mapbox GL JS (which now supports PMTiles natively as of v3.21), deck.gl, and more. You can switch formats with a single argument:

```{r}
#| eval: false
freestile(nc, "nc_mlt.pmtiles", layer_name = "counties", tile_format = "mlt")
```

### Ecosystem status

MLT was announced in January 2026, and library support is growing:

- **Encoding**: freestiler (Rust, R + Python), mlt-core reference encoder (Java/C++)
- **Decoding**: MapLibre GL JS (experimental, v5.21+), mlt-core decoder
- **Viewing**: MapLibre GL JS, mapgl in R
